---
title: ""   # Disable auto title slide
subtitle: ""
author: ""
date: ""
output:
  revealjs::revealjs_presentation:
    theme: white
    transition: slide
    slide_number: true
    self_contained: true
    center: true
    incremental: false
    toc: false
    number_sections: true
    toc_depth: 2
    code_folding: hide
    reveal_options:
      width: 1700
      height: 850
      minScale: 0.3
      maxScale: 1.0
    css: "style.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 8, fig.height = 5)
set.seed(801)

library(tidyverse)
library(MASS)
library(e1071)
library(class)
library(caret)
library(ggplot2)
library(gridExtra)

W_full <- read.csv("creditworthiness.csv")
names(W_full)[names(W_full) == "credit.rating"] <- "Creditworthy"
W_full$Creditworthy <- as.factor(W_full$Creditworthy)
W <- W_full %>% filter(Creditworthy != 0) %>% droplevels()
```

---

<!-- class: center, middle -->

<div style="font-size: 42px; font-weight: 700; margin-bottom: 20px;">
Support Vector Machines (SVM) for Credit Rating
</div>

<div style="font-size: 26px; color: #333; margin-bottom: 20px;">
STAT801 – Statistical Methods for Data Science (Enhanced)
</div>

<div style="font-size: 22px;">
Yeongjin Yu (5355436)  
`r format(Sys.Date(), '%d %b %Y')`
</div>

---

### 1. Introduction

- **Objective:** Build a multi-class classifier to predict customer **credit rating** (A, B, C = 1, 2, 3).  
- **Focus:** Support Vector Machines (SVM) — theory, tuning, and evaluation — benchmarked against **LDA, QDA, Naive Bayes, KNN**.  
- **Dataset:** 2 500 records, 46 predictors + 1 target (`Creditworthy`).

---

### 2. Data Overview

- **Labelled (1–3):** 1 962 records  
- **Unlabelled (0):** 538 records (to be classified after model selection)  
- **Predictors:** 46 variables – demographics, employment, savings, balances

```{r class_dist}
class_tbl <- as.data.frame(table(W$Creditworthy))
colnames(class_tbl) <- c("Class","Count")
ggplot(class_tbl, aes(x = Class, y = Count, fill = Class)) +
  geom_col() +
  geom_text(aes(label = Count), vjust = -0.3, size = 4) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none") +
  labs(title = "Labelled Class Distribution (n = 1,962)", y = "Count")
```
**Figure 1.** Labelled class distribution of creditworthiness (n = 1,962).

---

### 3. Data Preparation

- **Response:** `Creditworthy ∈ {1, 2, 3}`  
- **Scaling:** Applied for KNN & SVM  
- **Missing Values:** None  
- **Split (70/30):**

```{r split}
set.seed(801)
train_id <- sample(1:nrow(W), 0.7 * nrow(W))
train <- W[train_id, ];  test <- W[-train_id, ]
```

| Dataset | Samples | % |
|:--|--:|--:|
| Training | `r nrow(train)` | 70% |
| Test | `r nrow(test)` | 30% |

**Table 1.** Training and test dataset split summary.

---

### 4. Models Compared

1. **LDA** — Linear Discriminant Analysis  
2. **QDA (PCA)** — Quadratic Discriminant Analysis after PCA  
3. **Naive Bayes (NB)**  
4. **K-Nearest Neighbours (K = 5)**  
5. **SVM (RBF Kernel)** — tuned via grid search (C & γ)

---

### 5. LDA Results

```{r lda}
lda.fit <- lda(Creditworthy ~ ., data=train)
lda.pred <- predict(lda.fit, test)
lda.acc <- mean(lda.pred$class == test$Creditworthy)
```

- **Test Accuracy:** `r sprintf('%.2f', lda.acc)`  
- **Sensitivities:** Class 1 = 0.49, Class 2 = 0.75, Class 3 = 0.36  
- Linear boundaries favour major classes but underfit extremes.

---

### 6. QDA with PCA

```{r qda}
train.X <- scale(train[, -which(names(train) == "Creditworthy")])
test.X  <- scale(test[, -which(names(test) == "Creditworthy")])
pca.fit <- prcomp(train.X, center=TRUE, scale.=TRUE)
train.pca <- data.frame(pca.fit$x[,1:10], Creditworthy=train$Creditworthy)
test.pca  <- data.frame(predict(pca.fit,newdata=test.X)[,1:10], Creditworthy=test$Creditworthy)
qda.fit <- qda(Creditworthy~., data=train.pca)
qda.pred <- predict(qda.fit, test.pca)
qda.acc <- mean(qda.pred$class == test.pca$Creditworthy)
```

- **Test Accuracy:** `r sprintf('%.2f', qda.acc)`  
- First 10 PCs → ~30 % variance.  
- PCA stabilises numerics but reduces covariance information.

---

### 7. Naive Bayes & KNN

```{r nb_knn}
nb.fit <- naiveBayes(Creditworthy~., data=train)
nb.pred <- predict(nb.fit, test)
nb.acc <- mean(nb.pred == test$Creditworthy)
train.X2 <- scale(train[,-which(names(train)=="Creditworthy")])
test.X2  <- scale(test[,-which(names(test)=="Creditworthy")])
train.Y <- train$Creditworthy; test.Y <- test$Creditworthy
knn.pred <- knn(train.X2,test.X2,train.Y,k=5)
knn.acc <- mean(knn.pred == test.Y)
```

| Model | Accuracy | Note |
|:--|--:|:--|
| Naive Bayes | `r sprintf('%.2f', nb.acc)` | Strong independence assumption fails |
| KNN (k = 5) | `r sprintf('%.2f', knn.acc)` | High-dimensional sparsity reduces accuracy |
**Table 2.** Summary of Naive Bayes and KNN model performance.

---

### 8. SVM (RBF) Tuning

```{r svm_tune}
tune.out <- tune(svm, Creditworthy~., data=train,
                 kernel="radial",
                 ranges=list(cost=c(0.1,1,10,100),
                             gamma=c(0.01,0.1,1)))
best.model <- tune.out$best.model
best.pred <- predict(best.model, test)
best.acc <- mean(best.pred == test$Creditworthy)
perf <- tune.out$performances
perf$accuracy <- 1 - perf$error
```

- **Best Parameters:** C = 1, γ = 0.01  
- **Cross-validated Accuracy:** ≈ 0.61  
- **Test Accuracy:** `r sprintf('%.2f', best.acc)`  

```{r heatmap, fig.cap='Figure 2. Cross-validated accuracy by cost and gamma.'}
ggplot(perf, aes(x=factor(gamma), y=factor(cost), fill=accuracy)) +
  geom_tile(color="white") +
  scale_fill_viridis_c(option="C") +
  labs(x="Gamma (γ)", y="Cost (C)", fill="CV Accuracy") +
  theme_minimal(base_size=14)
```

---

### 9. Decision Boundary (2D Demo)

```{r decision, fig.cap='Figure 3. SVM Decision Boundary (2D Example).'}
df_plot <- train[, c(1,2,ncol(W))]
colnames(df_plot) <- c("X1","X2","Creditworthy")
svm.demo <- svm(Creditworthy~., data=df_plot, kernel="radial", cost=1, gamma=0.01)
grid <- expand.grid(X1=seq(min(df_plot$X1), max(df_plot$X1), length=100),
                    X2=seq(min(df_plot$X2), max(df_plot$X2), length=100))
grid$pred <- predict(svm.demo, grid)
ggplot(df_plot, aes(X1, X2, color=Creditworthy)) +
  geom_point(size=1.5) +
  geom_contour(data=grid, aes(z=as.numeric(pred)), color="black", bins=3) +
  theme_minimal(base_size=14)
```

---

### 10. Accuracy Comparison

```{r accuracy_plot}
acc_tbl <- tibble(
  Model=c("LDA","QDA (PCA)","Naive Bayes","KNN (k=5)","SVM (Best RBF)"),
  Accuracy=c(0.58,0.50,0.41,0.48,0.58))
ggplot(acc_tbl, aes(Model, Accuracy)) +
  geom_col() +
  geom_text(aes(label=sprintf('%.2f',Accuracy)),vjust=-0.3,size=4) +
  ylim(0,1) + theme_minimal(base_size=14)
```

---

### 11. Classifying 538 Unlabelled Cases


```{r predict_538}
unclassified <- W_full %>% filter(Creditworthy==0)
pred_unclassified <- predict(best.model,newdata=unclassified)
pred_dist <- as.data.frame(table(pred_unclassified))
colnames(pred_dist) <- c("Class","Count")
```

| Class | Count | % |
|:--|--:|--:|
| 1 | 28 | 5.2 % |
| 2 | 89 | 16.5 % |
| 3 | 421 | 78.3 % |

```{r dist_plot, fig.cap='Figure 5. Predicted distribution for 538 unlabelled cases.'}
ggplot(pred_dist, aes(Class, Count)) +
  geom_col(fill="#45B7D1") +
  geom_text(aes(label=Count), vjust=-0.3, size=4) +
  theme_minimal(base_size=14)
```

---

### 12. Conclusions

- **Best models:** SVM (RBF C = 1, γ = 0.01) & LDA ≈ 0.58 accuracy  
- **QDA (PCA):** 0.50 – stable but info loss (~30 % variance)  
- **NB & KNN:** lower due to assumptions and dimensionality  
- **Predictions:** 78 % of unlabelled → Class 3 (“Fair”)

**Takeaway:**  
A moderately complex SVM with small γ and moderate C best balances bias–variance and generalises well.
